Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-08 resubmission/documents for resubmission/Main_blinded.tex
Up carpeta resubmit

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic perceptions, and preferences for market justice
in Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with lower support for market justice; however, individuals
with strong meritocratic convictions are more likely to legitimize
existing disparities. The study also considers the influence of major
social movements during this period, which appear to have reshaped
public discourse and perceptions of fairness. These findings contribute
to a deeper understanding of how beliefs about justice and equity evolve
in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}

@@

Recent research has highlighted the importance of disaggregating the
concept of ``meritocratic beliefs,'' particularly by distinguishing
between meritocratic preferences and meritocratic perceptions
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
While preferences refer to normative ideals about how rewards should be
allocated (i.e., based on merit), perceptions refer to subjective
evaluations of how meritocracy actually operates in practice (i.e., how
rewards are allocated) (\citeproc{ref-janmaat_subjective_2013}{Janmaat,
2013}). This distinction allows for a more precise understanding of how
individuals make sense of inequality: people may endorse meritocracy
normatively while simultaneously recognizing its failure in practice. In
stratified societies, the gap between preferences and perceptions is
particularly evident, as individuals may continue to uphold the ideal of
meritocracy despite growing awareness of unequal opportunities
(\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).

@@

Meritocratic perceptions function as a normative framework that
legitimizes unequal outcomes, particularly when access to social goods
is stratified by income or socioeconomic background. Prior studies have
consistently shown that individuals who perceive society as meritocratic
express lower support for redistribution
(\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023};
\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}), greater legitimization of class inequality
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}), increased
tolerance for inequality (\citeproc{ref-day_movin_2017}{Day \& Fiske,
2017}), and lower perceptions of income inequality itself
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}). In the
regard of market justice, prior studies in Chile have shown that
individuals who perceive higher levels of meritocracy tend to express
stronger support for unequal distributions that reflect market outcomes
in healthcare, education and pensions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
Taken together, and as Busemeyer et al.
(\citeproc{ref-busemeyer_positive_2021}{2021}) argues, meritocratic
narratives can operate as feedback mechanisms that shape public opinion
and individual well-being by framing welfare outcomes as either deserved
or undeserved. This psychological mechanism underscores the normative
power of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal understandings of success and failure.

@@

Concerning public opinion, the ELSOC survey shows that Chile has a low
level of agreement with the idea that it is fair for those with higher
incomes to buy better welfare in the areas of education, healthcare, and
pensions, with average support of 8.5\% between 2016 and 2023. At the
same time, around 55\% and 47\% of the public between 2016 and 2023
\emph{disagreed} that effort and talent are rewarded accordingly. This
coincides with 90\% of citizens \emph{agreeing} that income differences
in Chile are \emph{too large}
(\citeproc{ref-castillo_changes_2025}{Castillo et al., 2025}).

In sum, Chile represents an illustrative case of a particular form of
\emph{moral economy} (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
2019}), in which national socioeconomic characteristics, institutional
arrangements, and public attitudes coexist. This suggests that the
country combines persistently high levels of economic and social
inequality with a strong institutionalization of market-based mechanisms
within the welfare system
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). At the same
time, perceptions of inequality are high, and there is widespread public
criticism of the commodification of social services
(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). For the above
reasons, we argue that the Chilean case represents a relevant context to
study the role of perceived income inequality and meritocracy in shaping
public preferences regarding the commodification of welfare services,
particularly from the perspective of individual change.

@@

Building upon the previous literature, this study proposes that
attitudes toward market justice are shaped by a dynamic interplay
between individuals' perceptions of economic inequality and their
perceptions of meritocracy. Specifically, we argue that both perceptions
independently and interactively influence the extent to which
individuals endorse market-based distributions of social goods and
services in Chile.

First, consistent with previous findings about redistributive
preferences (\citeproc{ref-castillo_perception_2022}{Castillo et al.,
2022}), we expect that a higher perception of economic inequality will
be associated with lower market justice preferences. When individuals
perceive smaller income gaps, they are more likely to view market
mechanisms as fair and legitimate, reinforcing the acceptance of
outcomes based on competition and ability to pay. Conversely, a
heightened perception of inequality may erode confidence in market
fairness, weakening support for market-based distribution. This
relationship is particularly relevant in the context of Chile, where the
neoliberal economic model has been a dominant force in shaping public
attitudes toward inequality and justice
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021};
\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}).

@@

Additionally, this study examines whether these relationships vary over
time, particularly in light of broader social and political
transformations. We are especially interested in whether changes in
individuals' perceptions of inequality and meritocracy are associated
with shifts in their market justice attitudes across different time
points. While we acknowledge the relevance of contextual events --- such
as the 2019 political outburst or the subsequent constitutional process
--- for public opinion and political orientation
(\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
2025}), identifying causal effects of such events (e.g., mass protest
exposure) on individual attitudes lies outside the scope of this
research. Instead, we focus on how intra-individual trajectories in
perceived inequality and merit shape evolving attitudes toward the
commodification of social services.

@@

This study tests several key hypotheses concerning the relationships
between perceptions of economic inequality, meritocratic perceptions,
and preferences toward market justice in Chile. It posits that higher
perceived inequality and stronger meritocratic perceptions both
independently contribute to greater support for market-based
distributions of social goods. Furthermore, the study examines how
perceived economic inequality moderates the influence of meritocratic
perceptions on market justice preferences. Additionally, it considers
how major social and political mobilizations, such as those occurring
around 2019 and subsequent years, may have weakened these relationships,
reflecting increased critical perspectives on systemic inequalities and
market mechanisms. The subsequent discussion interprets the empirical
findings in light of these hypotheses, considering their implications
for understanding attitudes toward inequality and social justice in a
context marked by profound social change.

The first set of hypothesis suggested that perceptions of economic
inequality were strongly linked to preferences for market justice. The
findings demonstrated that individuals perceiving higher income
disparities showed less support for the idea that the access to core
social services depends on individual income, aligning with theories
emphasizing that awareness of inequality could fuel a critical view of
such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
reflects a moral economy in which perceptions of systemic unfairness
undermine the legitimacy of existing distributions and fuel demands for
greater equity. In particular, the negative within-individual effect
observed over time (\(H1_{b}\)) suggests that as inequality becomes more
salient for some individuals, their support for market justice
declines---potentially due to rising distrust in market mechanisms or
growing disillusionment with the fairness of the system. However,
although the effect of perceived inequality at the between-individual
level is also negative, it is not statistically significant. Therefore,
we cannot conclude that individuals who, on average, perceive higher
inequality are systematically less supportive of market justice. As a
result, hypothesis \(H1_{a}\) does not receive empirical support.

Regarding the second set of hypotheses, the results confirmed that
meritocratic perceptions---particularly those emphasizing individual
effort---were associated with stronger support for market-based
distribution systems. Individuals who believed that success is primarily
achieved through personal effort were more likely to justify unequal
access to core social services based on income, interpreting such
disparities as outcomes of individual merit rather than systemic
injustice (\(H2_{a}\)). This finding aligns with previous research
showing that meritocratic narratives serve as moral justifications that
legitimize social stratification
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
operate symbolically to reinforce structural inequalities by reducing
support for redistributive policies, framing inequality as both fair and
deserved. Consistent with prior work by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
the findings suggest that such meritocratic perceptions uphold existing
hierarchies by promoting the acceptance of inequality as a reflection of
individual virtue rather than structural failure. This mechanism is
particularly salient in neoliberal contexts like Chile, where market
logics heavily shape social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

Interestingly, intra-individual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
that rewards are based on individual effort are associated with stronger
preferences for market justice, increases in the belief that rewards
derive from talent are linked to weaker support for such principles. One
possible explanation is that effort is generally viewed as a
controllable and malleable trait, aligning with the normative
foundations of market fairness. In contrast, talent tends to be
perceived as innate and less subject to personal control, rendering
talent-based inequality less legitimate. Furthermore, increased exposure
to real-world scenarios in which outcomes are clearly shaped by inherent
traits rather than hard work may lead individuals to question the
fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions
of inequality in the relationship between meritocratic perceptions and
preferences for market justice. The analysis revealed that the positive
association between meritocratic perceptions and support for
market-based distribution systems tends to be stronger when perceived
economic inequality is high---contradicting our initial expectations
(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
statistically significant for effort-related meritocratic perceptions,
it is significant for talent-related perceptions at both the within- and
between-individual levels. This finding runs counter to our hypotheses,
as it suggests that the effect of meritocratic
perceptions---specifically, talent-based meritocracy---intensifies as
perceptions of inequality increase. However, this interpretation
requires caution. The main effect of perceived inequality on support for
market justice is negative. Thus, as perceived inequality becomes less
negative (i.e., closer to zero), the positive relationship between
talent-based meritocratic perceptions and market justice preferences
becomes stronger. This suggests a more nuanced dynamic: meritocratic
perceptions, particularly those emphasizing talent, may serve as a
stronger justificatory mechanism for market-based inequalities among
individuals who are less inclined to perceive inequality as a problem.
In other words, among those who do not strongly perceive systemic
disparities, meritocratic narratives may play a more influential role in
legitimizing unequal outcomes.

Finally, the fourth hypothesis examined the potential impact of recent
social mobilizations---particularly those that erupted in 2019---on the
relationship between meritocratic perceptions, perceptions of
inequality, and preferences for market justice. The main effect of time
indicates that, on average, support for market justice increased
following the 2019 protests. However, the analysis did not detect any
significant interactions between distributive beliefs and support for
market justice, suggesting that the social mobilizations did not
substantially modify the underlying associations between these
variables. As a result, hypotheses \(H4_{a}\) and \(H4_{b}\) are not
supported by the data. This may point to a certain stability---or even
resilience---in the normative frameworks that guide individuals'
evaluations of distributive justice, despite the occurrence of major
collective political events. Alternatively, while the protests may have
eroded trust in institutional arrangements or governance, they may not
have fundamentally altered individuals' beliefs about how rewards and
resources should be allocated in society.

@@

This study examined the complex interplay between perceptions of
economic inequality, meritocratic perceptions, and preferences toward
market justice in Chile from 2016 to 2023, drawing on longitudinal data
from the ELSOC survey. By exploring how subjective assessments and
social contexts influence support for redistribution and market-based
resource allocation, the research offers different elements that
contribute to the understanding of the normative foundations
underpinning social justice attitudes in a highly unequal and
commoditized environment.

The findings confirm that higher perceptions of economic inequality are
associated with less support for market justice preferences, this is,
the belief that it is fair that those with higher income have better
social services such as education, pensions and health. At the same
time, meritocratic perceptions---particularly those emphasizing
individual effort---are strongly associated with support for
market-based distributions, suggesting that meritocracy serve as a moral
justification for structural inequalities. However, changes in
meritocratic perceptions over time reveal a more nuanced picture: while
increased emphasis on effort reinforces support for market justice,
increased emphasis on talent tends to reduce it. This distinction likely
reflects broader beliefs about the controllability and fairness of
different meritocratic traits. Moreover, the interaction between
inequality perceptions and meritocratic perceptions indicates that the
legitimizing power of meritocracy becomes stronger as perceptions of
inequality become less negative---that is, when individuals are less
aware or concerned about economic disparities. This finding highlights a
potential feedback mechanism by which lower sensitivity to inequality
may enable stronger endorsement of merit-based explanations for unequal
outcomes.

This research advances the extant literature by integrating subjective
perceptions with social and political contexts to explain attitudes
toward economic inequality and distributional justice. While previous
studies primarily focused on objective measures or individual
characteristics (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}), this work emphasizes the
dynamic and interactional nature of perceptions and beliefs over time.
It highlights the importance of socio-political upheavals in reshaping
normative attitudes, underscoring the role of collective action in
challenging entrenched narratives of meritocracy and fairness. The
longitudinal approach provides a deeper temporal perspective on how
societal events could influence individual perceptions and preferences.

Despite its contributions, the study has several limitations. First, the
reliance on self-reported perceptions and attitudes may be subject to
social desirability bias. Second, the focus on Chile, while offering
valuable insights into a specific context, limits the generalizability
of results to other countries with different institutional and cultural
backgrounds. Third, the observational nature of the data prevents causal
inferences, and unobserved confounding factors may influence the
observed relationships. Finally, the potential influence of media,
political discourse, and peer effects on perceptions was not directly
assessed.

Future research could explore the causal mechanisms linking perceptions
and attitudes, possibly through experimental or mixed-method designs.
Comparative studies across different national contexts would help
determine the universality or specificity of these dynamics.
Additionally, investigating the role of media, political communication,
and education in shaping perceptions of inequality and meritocracy would
deepen understanding of the normative foundations of social justice
attitudes. Finally, examining how these perceptions influence behavioral
outcomes, such as political participation or support for social
movements, would provide valuable insights into the pathways from
beliefs to collective action and policy change.

@@

\bibitem[\citeproctext]{ref-castillo_changes_2025}
Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
{Justification} of {Educational Inequalities}: {The Role} of
{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
Pandemic}. \emph{Social Justice Research}.
\url{https://doi.org/10.1007/s11211-025-00458-0}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}
\author{ }

\date{}
\begin{document}
\maketitle
\begin{abstract}
Several countries have experienced a shift toward the privatization and
commodification of public goods, welfare policies, and social services.
In Latin America, Chile stands out as a paradigmatic case where this
trend has led to the extensive marketization of essential services. From
a moral economy perspective, the extent to which individuals consider it
fair for access to such services to depend on market criteria has been
conceptualized as market justice preferences. This study investigates
the relationship between perceptions of economic inequality,
meritocratic beliefs, and market justice preferences in Chile between
2016 and 2023. Using six waves of panel data from the Chilean
Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
\(N_{individuals}\) = 1,687), the analysis examines how subjective
assessments of inequality shape attitudes toward the role of merit in
access to key social services such as healthcare, education, and
pensions. The findings show that greater perceived inequality is
associated with lower market justice preferences. However, individuals
who believe that effort is rewarded are more likely to legitimize
existing disparities. In contrast, the perception that talent is
rewarded shows a negative effect on market justice preferences; an
effect that intensifies as perceived inequality increases over time. The
study also considers the influence of major social movements during this
period, which appear to have reshaped public discourse on justice and
fairness. These findings contribute to a deeper understanding of how
distributive beliefs evolve in contexts marked by persistent inequality
and entrenched neoliberal frameworks \newline \textbf{Keywords}:
Inequality perception, meritocracy, commodification, Chile, public
preferences
\end{abstract}

@@

Recent research has highlighted the importance of disentangling the
concept of ``meritocratic beliefs,'' particularly by distinguishing
between meritocratic preferences and meritocratic perceptions
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
While preferences refer to normative ideals about how rewards should be
allocated (i.e., based on merit), perceptions refer to subjective
evaluations of how meritocracy actually operates in practice (i.e., how
rewards are allocated) (\citeproc{ref-janmaat_subjective_2013}{Janmaat,
2013}). This distinction allows for a more precise understanding of how
individuals make sense of inequality: people may endorse meritocracy
normatively while simultaneously recognizing its failure in practice. In
stratified societies, the gap between preferences and perceptions is
particularly evident, as individuals may continue to uphold the ideal of
meritocracy despite growing awareness of unequal opportunities
(\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).

@@

Meritocratic perceptions function as a normative framework that
legitimizes unequal outcomes, particularly when access to social goods
is stratified by income or socioeconomic background. Prior studies have
consistently shown that individuals who perceive society as meritocratic
express lower support for redistribution
(\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023};
\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
2025}), greater legitimization of class inequality
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}), increased
tolerance for inequality (\citeproc{ref-day_movin_2017}{Day \& Fiske,
2017}), and lower perceptions of income inequality itself
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
Regarding market justice preferences, prior studies in Chile have shown
that individuals who perceive higher levels of meritocracy tend to
express stronger support for unequal distributions that reflect market
outcomes in healthcare, education and pensions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
Taken together, and as Busemeyer et al.
(\citeproc{ref-busemeyer_positive_2021}{2021}) argues, meritocratic
narratives can operate as feedback mechanisms that shape public opinion
and individual well-being by framing welfare outcomes as either deserved
or undeserved. Such mechanism underscores the normative power of
meritocracy in stabilizing unequal systems by shaping political
attitudes and personal understandings of success and failure.

@@

@@

Building upon the previous literature, this study proposes that
attitudes toward market justice are shaped by a dynamic interplay
between individuals' perceptions of economic inequality and their
perceptions of meritocracy. Specifically, we argue that both perceptions
independently and interactively influence the extent to which
individuals endorse market-based distributions of social goods and
services in Chile. To our knowledge, there are not studies so fay
analyzing such associations, let alone from a longitudinal perspective.

Firstly, and consistent with previous findings about research on
redistributive preferences
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}), we
expect that a higher perception of economic inequality will be
associated with lower market justice preferences. When individuals
perceive smaller income gaps, they are more likely to view market
mechanisms as fair and legitimate, reinforcing the acceptance of
outcomes based on competition and ability to pay. Conversely, a
heightened perception of inequality may erode confidence in market
fairness, weakening support for market-based distribution. This
relationship is particularly relevant in the context of Chile, where the
neoliberal economic model has been a dominant force in shaping public
attitudes toward inequality and justice
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021};
\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}).

@@

Additionally, this study examines whether these relationships vary over
time, particularly in light of broader social and political
transformations. We are especially interested in whether changes in
individuals' perceptions of inequality and meritocracy are associated
with shifts in their market justice attitudes across different time
points. While we acknowledge the relevance of contextual events for
public opinion and political orientation
(\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
2025}), such as the 2019 political outburst or the subsequent
constitutional process in CHile, we are aware that identifying causal
associations lies outside the scope of this research. Instead, and from
an exploratory perspective, we focus on how intra-individual
trajectories in perceived inequality and merit shape evolving attitudes
toward the commodification of social services.

@@

The first set of hypotheses proposed that perceptions of economic
inequality are strongly associated with preferences for market justice.
In the between-individual specification, the effect of perceived
inequality is negative but not statistically significant. Therefore, we
cannot conclude that individuals who, on average, perceive higher levels
of inequality are systematically less supportive of market justice. As a
result, hypothesis \(H1_{a}\) is not supported by the data. However, in
the within-individual specification, the findings reveal a significant
negative association: individuals who, at a given point in time,
perceive higher income disparities are less supportive of the idea that
access to core social services should depend on individual income. This
result aligns with theories suggesting that increased awareness of
inequality fosters a more critical stance toward market-based
distributive arrangements
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a pattern reflects a
moral economy logic in which perceptions of systemic unfairness
undermine the legitimacy of existing distributions and strengthen
demands for greater equity. Specifically, the negative within-individual
effect over time (\(H1_{b}\)) suggests that as inequality becomes more
salient for some individuals, their support for market justice
declines---possibly due to growing distrust in market mechanisms or
increasing disillusionment with the perceived fairness of the system.

Regarding the second set of hypotheses, the results confirmed that
meritocratic perceptions---particularly those emphasizing individual
effort---were associated with stronger support for market-based
distribution systems. Individuals who believed that success is primarily
achieved through personal effort were more likely to justify unequal
access to core social services based on income, interpreting such
disparities as outcomes of individual merit rather than systemic
injustice (\(H2_{a}\)). This finding aligns with previous research
showing that meritocratic narratives serve as moral justifications that
legitimize social stratification
(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
operate symbolically to reinforce structural inequalities by reducing
support for redistributive policies, framing inequality as both fair and
deserved. Consistent with prior work by Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
the results suggest that such meritocratic perceptions uphold existing
hierarchies by promoting the acceptance of inequality as a reflection of
individual virtue rather than structural failure. This mechanism is
particularly salient in neoliberal contexts like Chile, where market
logics heavily shape social attitudes
(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

Interestingly, intra-individual changes in meritocratic perceptions over
time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
that rewards are based on individual effort are associated with stronger
preferences for market justice, increases in the belief that rewards
derive from talent are linked to weaker support for such principles. One
possible explanation is that effort is generally viewed as a
controllable and malleable trait, whereas talent tends to be perceived
as innate and less subject to personal control, rendering talent-based
inequality less legitimate. Furthermore, increased exposure to
real-world scenarios in which outcomes are clearly shaped by inherent
traits rather than hard work may lead individuals to question the
fairness and legitimacy of market-based reward systems.

The third set of hypotheses addressed the moderating role of perceptions
of inequality in the relationship between meritocratic perceptions and
preferences for market justice. The analysis revealed that the positive
association between meritocratic perceptions and support for
market-based distribution systems tends to be stronger when perceived
economic inequality is high, contradicting our initial expectations
(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
statistically significant for effort-related meritocratic perceptions,
it is significant for talent-related perceptions at both the within- and
between-individual levels. However, this interpretation requires
caution: the main effect of perceived inequality on support for market
justice is negative, meaning that as perceived inequality becomes less
negative (i.e., closer to zero), the positive relationship between
talent-based meritocratic perceptions and market justice preferences
becomes stronger. This suggests a more nuanced dynamic: meritocratic
perceptions, particularly those emphasizing talent, may serve as a
stronger justificatory mechanism for market-based inequalities among
individuals who are less inclined to perceive inequality as a problem.
In other words, among those who do not strongly perceive systemic
disparities, meritocratic narratives may play a more influential role in
legitimizing unequal outcomes.

Finally, the fourth exploratory hypothesis examined the potential impact
of social events --- particularly those that erupted in 2019 --- on the
relationship between meritocratic perceptions, perceptions of
inequality, and preferences for market justice. The main effect of time
indicates that, on average, support for market justice increased after
2019 protests. However, the analysis did not detect any significant
interactions between distributive beliefs and support for market
justice, suggesting that events in this period (as the mobilizations,
the pandemic and/or the constitucional processes) did not substantially
modify the underlying associations between these variables. As a result,
hypotheses \(H4_{a}\) and \(H4_{b}\) are not supported by the analysis.
This may point to a certain stability or even resilience in the
normative frameworks that guide individuals' evaluations of distributive
justice, despite the occurrence of major collective political events.
Alternatively, while the protests may have eroded trust in institutional
arrangements or governance, they may not have fundamentally altered
individuals' beliefs about how rewards and resources should be allocated
in society.

@@

This study examined the complex interplay between perceptions of
economic inequality, meritocratic perceptions, and preferences toward
market justice in Chile from 2016 to 2023, drawing on longitudinal data
from the ELSOC survey. By exploring how subjective assessments and
social contexts influence support for redistribution and market-based
resource allocation, the research offers different elements that
contribute to the understanding of the normative foundations
underpinning social justice attitudes in a highly unequal and
commodified environment.

The findings support that higher perceptions of economic inequality are
associated with less support for market justice preferences, this is,
the belief that it is fair that those with higher income have better
social services such as education, pensions and health. At the same
time, meritocratic perceptions --- particularly those emphasizing
individual effort --- are strongly associated with support for
market-based distributions, suggesting that meritocracy serve as a moral
justification for structural inequalities. However, changes in
meritocratic perceptions over time reveal a more nuanced picture: while
increased emphasis on effort reinforces support for market justice,
increased emphasis on talent tends to reduce it. This distinction likely
reflects broader beliefs about the controllability and fairness of
different meritocratic traits. Moreover, the interaction between
inequality perceptions and meritocratic perceptions indicates that the
legitimizing power of meritocracy becomes stronger as perceptions of
inequality become less negative --- that is, when individuals are less
aware or concerned about economic disparities. Such finding highlights a
potential feedback mechanism by which lower sensitivity to inequality
may enable stronger endorsement of merit-based explanations for unequal
outcomes.

This research advances the extant literature by integrating subjective
perceptions with social and political contexts to explain attitudes
toward economic inequality and distributional justice. While previous
studies primarily focused on objective measures or individual
characteristics (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}), this work emphasizes the
dynamic and interactional nature of perceptions and beliefs over time.
Furthermore, it highlights the importance of socio-political upheavals
in reshaping normative attitudes, underscoring the role of collective
action in challenging entrenched narratives of meritocracy and fairness.
The longitudinal approach provides a deeper temporal perspective on how
societal events could influence individual perceptions and preferences.

Regarding avenues for future research, experimental evidence could help
expand the understanding of the association between perceptual variables
and market justice preferences. International comparative studies are
also necessary in order to assess the role of national contexts and the
universality or specificity of these dynamics. Additionally,
investigating the role of media, political communication, and education
in shaping perceptions of inequality and meritocracy would deepen
understanding of the normative foundations of social justice attitudes
and their variation over time. Finally, examining how these perceptions
influence behavioral outcomes, such as political participation or
support for social movements, would provide valuable insights into the
pathways from beliefs to collective action and policy change.

@@

### 2025-08-08 resubmission/paper-blinded-resubmit.qmd
Up carpeta resubmit

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra-Sanhueza
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
citeproc: true
abstract: | 
 Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services. In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services. From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences. This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. The findings show that greater perceived inequality is associated with lower market justice preferences. However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities. In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness. These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
  \newline
 **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
